Equalizer\\Programming and User Guide

### 2009-07-31 Developer/ProgrammingGuide/paper.tex
Release Candidate of Programming Guide
[ ] Will break Win32 build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications based on
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.5.10 for Equalizer 0.6+svn\\[\medskipamount]
%  \today
}

@@

\lstset{language=C++}
\section*{Equalizer 0.6 Programming and User Guide}
\today
\subsection*{Contributors}

@@

OpenGL is a registered trademark, OpenGL Multipipe is a trademark of
Silicon Graphics, Inc. Linux is a registered trademark of Linus
Torvalds.  Mac OS is a trademark of Apple Inc. CAVELib is a registered
trademark of the University of Illinois. The CAVE is a registered
trademark of the Board of Trustees of the University of Illinois at
Chicago. Qt is a registered trademark of Trolltech. All other
trademarks and copyrights herein are the property of their respective
owners.

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
    1.5.4   & Feb 09, 2009 & Added \sref{sNetUsage}\\
    1.5.5   & Mar 10, 2009 & Updated to reflect new AGL event dispatch\\
    1.5.6   & Mar 23, 2009 & Documented Layout API\\
    1.5.7   & Apr 17, 2009 & Updated Layout API (multiple observers)\\
    1.5.8   & May 30, 2009 & Added \sref{sEqualizers}, updated \sref{sLoadBalancing} and \sref{sSwapBarrier}\\
    1.5.9   & Jul 27, 2009 & Major review pass before release\\
    1.5.10  & Jul 28, 2009 & Split in User and Programming Guide, added \sref{sNamespaces}\\
    \hline
  \end{tabularx}}
\clearpage

@@

Equalizer is the next step in the evolution of middleware for scalable
OpenGL applications. Existing solutions, such as OpenGL Multipipe SDK,
Cavelib and VRJuggler, implement a subset of Equalizer. In other areas,
e.g., tracking device support, some of them provide more functionality.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
  {\caption{\label{fExecutionFlow}Parallel Rendering}}
  \vspace{-3ex}
\end{wrapfigure}

\fig{fExecutionFlow} illustrates the basic principle of any parallel
rendering application. The typical OpenGL application, for example
GLUT-based applications, has an event loop which redraws the scene,
updates data based on received events, and eventually redraws a new
frame.

A parallel rendering application uses the same basic execution model and
extends it by separating the rendering code from the main event
loop. The rendering code is then executed in parallel on different
resources. This model is naturally followed by Equalizer, thus making
application development as easy as possible.

@@

The Equalizer architecture is based on a client-server model. The client
library exposes all functionality discussed in this document to the
programmer, and provides communication between the different Equalizer
processes.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/processes.pdf}
  {\caption{\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The render client implements the rendering part of an application. Its
execution is passive, it has no main loop and is completely driven by
Equalizer, based on the rendering tasks received from the server. The
tasks are executed by calling the appropriate task methods (see
\sref{sTaskMethods}) in the correct thread and context. The application
either implements the task methods with application-specific code or
uses the default methods provided by Equalizer.

The application can also be a rendering client, in which case it can
also contribute to the rendering. If it does not implement any render
client code, it is reduced to be the application's `master' process
without any OpenGL windows and rendering code.

@@

This section gives an introduction to scalable rendering, to
provide some background for application developers. The scalability
modes offered by Equalizer are discussed, along with their advantages
and disadvantages.

@@

2D decompositions can be used by all types of applications, but should
be combined with DB compounds to reduce the data per node, if
possible. In most cases, a \textsf{loadBalancer} should be used to
automatically adapt the tiling each frame, based on the current
rendering load.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\label{fPixel}\small A pixel compound}}
\end{wrapfigure}
OpenGL functionality influenced by the raster position will not work
correctly with pixel compounds, or needs at least special
attention. Among them are: lines, points, sprites, glDrawPixels,
glBitmap, glPolygonStipple. The application can query the current pixel
parameters at run-time to adjust the rendering accordingly.

Pixel compounds work well for purely fill-limited applications.
Techniques like frustum culling do not reduce the rendered data for the
source rendering resources. Pixel compounds are ideal for raytracing,
which is highly fill-limited and needs the full database for rendering
anyway. Volume rendering applications are also well suited for this
mode, and should choose it over 2D compounds.

@@

Some scalable rendering parameters are updated at run-time to provide
optimal performance or influence other scalable rendering
features. These adjustments are often referred to as load-balancing, but
are called \textsf{equalizers}, since their functionality is not limited
to load-balancing in Equalizer.

The server provides a number of equalizers, which automatically update
certain parameters of compounds based on run-time information. They
balance the load of 2D and DB compounds, optimally distribute render
resources for multi-segment projections systems, adapt the resolution to
provide a constant framerate or zoom images to allow monitoring of
another view.

@@

\begin{itemize}
\item A list of all \textbf{computers} in you rendering cluster,
  including the \textbf{IP addresses} of all network interfaces to be
  used.
\item The number of \textbf{graphics cards} in each computer.
\item The \textbf{physical dimensions} of the display system, if
  applicable. These are typically the bottom-left, bottom-right and
  top-left corner points of each display surfaces in meters.
\item The relative coordinates of all the \textbf{segments} belonging to
  each display surface, and the graphics card \textbf{output} used for
  each segment. For homogenous setups, it is often enough to know the
  number of rows and columns on each surface, as well as the overlap or
  underlap percentage, if applicable.
\item The number of desired \textbf{application windows}. Application
  windows are typically destination windows for scalable rendering or
  windows paired with a view on a display system.
\item \textbf{Characteristics} of the application, e.g., supported
  scalability modes and features.
\end{itemize}

@@

Equalizer applications are configured at run-time by the Equalizer
server. The server loads its configuration from a text file, which is an
one-to-one representation of the configuration data structures at
runtime.

@@

In each node, list all \textsf{connection}s thorugh which this node is
reachable. Typically a node uses only one connection, but it is possible
to configure multiple connections if the machine and cluster is set up
to use multiple, independent network interfaces. Make sure the
\textsf{hostname} used is reachable from all nodes. An IP address may be
used as the hostname.

@@

For each node, create on \textsf{pipe} for each graphics card in the
machine. Set the \textsf{device} number to the correct index. On
operating systems using X11, e.g., Linux, also set the port number if
your X-Server is running on a non-standard port.

The \textsf{eq::Pipe} class is the abstraction of a graphics card (GPU).
In the current implementation it is also one operating system
thread. Non-threaded pipes are supported for integrating with
thread-unsafe libraries, but have various performance caveats. They
should only be used if using a different, but synchronized rendering
thread is not an option.

@@

Configure one \textsf{window} for each desired application window on the
appNode. Configure one full-screen window for each display
segment. Configure one off-screen window, typically a pbuffer, for each
graphics card used as a source for scalable rendering.

Sometimes display segments cover only a part of the graphics card
output. In this case it is adviced to configure a non-fullscreen window
without window decorations, using the correct window viewport.

@@

Configure one \textsf{channel} for each desired rendering in each window
area. Typically one full-screen channel per window is used. Name the
channel using a unique, easily-identifiable name.

@@

Sometimes, a single window is split across multiple projectors, e.g., by
using an external splitter such as the Matrox triple-head-to-go. In this
case configure one channel for each segment, using the channel's
\textsf{viewport} to configure its position wit respect to the window.

@@

A canvas represents one logical projection surface, e.g., a powerwall, a
curved screen or an immersive installation. One configuration might
drive multiple canvases, for example an immersive installation and an
operator station.

A canvas consists of one or more segments. A planar canvas typically has
a frustum description (see \sref{sFrustum}), which is inherited by the
segments. Non-planar frusta are configured using the segment
frusta. These frusta typically describe a physically-correct display
setup for Virtual Reality installations.

@@

For non-planar displays, configure the frustum as described in
\sref{sCanvas}. For passive stereo installations, configure one segment
per eye pass, where the segment for the left and right eye fully
have the same viewport.

A segment represents one output channel of the canvas, e.g., a projector
or a LCD. A segment has an output channel, which references the channel to
which the display device is connected.

A segment covers a part of its parent canvas, which is configured using
the segment viewport. The viewport is in normalized coordinates with
respect to the canvas. Segments might overlap (edge-blended projectors)
or have gaps between each other (\fig{fCanvas}\footnote{Dataset courtesy
  of VolVis distribution of SUNY Stony Brook, NY, USA.}). The viewport
is used to configure the segment's default frustum from the canvas
frustum description, and to place layout views correctly.

@@

A view has a fractional viewport with respect to its layout.  A layout
is often fully covered by its views, but this is not a requirement.

@@

Compound trees are used to describe how the rendering resources are
combined to produce the desired output, and how they are aggregated to
increase the performance. It is adviced to study and understand the
basic configuration files shipped with the Equalizer configuration,
before attempting to write compound configurations.

For configurations using canvases and layouts without scalability, the
configuration file loader will create the appropriate compounds.

Usage of the rendering resources is configured using a compound
tree. Although the API does not currently expose compounds, the basic
design behind compounds is explained here for a better understanding of
the Equalizer architecture. Further information on the configuration of
compounds can be found on the Equalizer website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/compounds.html}}.

@@

Compounds have a frustum description to define the physical layout of
the display environment.

@@

The frustum can be specified as described in \sref{sCanvas}.

@@

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination channel's rendering, which is defined by the viewport,
frustum and database. A \textsf{viewport} decomposes the destination
channel and frustum in screen space. A \textsf{range} tells the
application to render a part of its database, and an \textsf{eye}
rendering pass can selectively render different stereo passes. A
\textsf{pixel} parameter adapts the frustum so that the source channel
renders an evenly subset of the parent's pixels. Setting one or multiple
attributes causes the parent's view to be decomposed
accordingly. Attributes are cumulative, that is, intermediate compound
attributes affect and therefore decompose the rendering of all their
children.

@@

Note that a software swap barrier is not accurate enough for
edge-blended projection system. Such installations typically use a
hardware synchronization, e.g., nVidia G-Sync cards. It is however
sufficient for display walls made out of LCD's and for some
non-immersive projection systems.

@@

\subsubsection{\label{sEqualizers}Adaptation - Equalizers}

Equalizers are used to update compound parameters based on run-time
data. They are attached to a compound (sub-)tree, on which they
operate. The Equalizer distribution contains numerous example
configuration files using equalizers.

@@

Using a 2D or DB load-balancer will adapt the 2D split or database range
automatically each frame. The 2D load-balancer exists in three flavors:
\textsf{2D} using tiles, \textsf{VERTICAL} using columns and
\textsf{HORIZONTAL} using rows.

@@

DB load-balancing is beneficial for applications which cannot precisely
predict the load for their scene data, e.g., when the data is
non-uniform. Volume rendering is a counterexample, where the data is
uniform and a static DB decomposition typically results in a better
performance.

\paragraph{View Equalizer}
Depending on the model position and data structure, each segment of a
multi-display system has a different rendering load. The segment with
the biggest load determines the overall performance when using a static
assignment of resources to segments. The view equalizer analyzes the
load of all segments, and adapts the resource usage each frame. It
equalizes the load on all segments of a view.

\fig{fViewLoadBalancing}\footnote{Image Copyright Realtime Technology
  AG, 2008} illustrates this process. On the left side, a static
assigment of resources to display segments is used. The right-hand
segment has a higher load than the left-hand segment, causing
sub-optimal performance. The configuration on the left uses a view
equalizer, which assigns two GPU's to the left segment and four GPU's to
the right segment, which leads to optimal performance for this model and
camera position.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
  {\caption{\label{fViewLoadBalancing}\small Cross-segment
      load-balancing for two segments using eight GPU's}}
\end{wrapfigure}
The view equalizer can also use resources from another display resource,
if this resource has little rendering load by itself. It is therefore
possible to improve the rendering performance of a multi-display system
without any additional resources. This is particulary useful for
installations with a higher number of displays where the rendering load
is typically in a few segments only, e.g., for a CAVE.

@@

\paragraph{DFR Equalizer}
\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
  {\caption{\label{fDFR}\small Dynamic Frame Resolution}}
\end{wrapfigure}
Dynamic Frame Resolution (DFR) trades rendering performance for visual
quality. The rendering for a channel is done at a different resolution
than the native channel resolution to keep the framerate constant. The
DFR equalizer adapts the zoom of a channel, based on the target and
current framerate. It is typically used for fill-rate bound
applications, such as volume rendering and raytracing.

\fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
  Science Institute, University of Freiburg, Germany} shows DFR for
volume rendering. To achieve 10 frames per second, the model is rendered
at a lower resolution, and upscaled to the native resolution for
display. The rendering quality is slightly degraded, while the rendering
performance remains interactive. When the application is idle, it
renders a full-resolution view.

The dynamic frame resolution is not limited to down-scaling the
rendering resolution, it will also up-scale the resolution if the
parameters allow for it. Upscaled rendering, which will downsample the
result for display, provides dynamic anti-aliasing at a constant
framerate.

\paragraph{Monitor Equalizer}
The monitor equalizer allows the observation of another view,
potentially made of multiple segments, in a different channel at a
different resolution. This is typically used to re-use the rendering of
a large-scale display on an operator station.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/monitorEq.pdf}
  {\caption{\label{fMonitorEq}\small Monitoring a projection wall}}
\end{wrapfigure}
The monitor equalizer adapts the frame zoom of the output frames used to
observe the rendering, depending on the destination channel size. The
output frames are downscaled on the GPU before readback, which results
in optimal performance.

@@

To adapt an application for Equalizer, the programmer structures the
source code so that the OpenGL rendering can be executed in parallel,
potentially using multiple processes for cluster-based execution.

@@

Methods called by the application have the form \textsf{verb[Noun]},
whereas methods called by Equalizer (`Task Methods') have the form
\textsf{nounVerb}. For example, the application calls
\textsf{Config::startFrame} to render a new frame, which causes --among
other things-- \textsf{Node::frameStart} to be called in all active
render clients.

@@

\begin{description}
\item[eq::base] The eq::base namespace provides C++ classes to abstract
  the underlying operating system and implements common helper
  functionality. One example is the \textsf{eq::base::Clock} class
  providing a high-resolution timer. Classes in this namespace are well
  documented on the API documentation on the Equalizer website, and are
  not subject of this Programming Guide.
\item[eq::net] The Equalizer network layer provides basic functionality
  for network communication, such as \textsf{Connection} and
  \textsf{ConnectionSet}, as well as higher-level functionality such as
  \textsf{Node}, \textsf{Session} and \textsf{Object}. Please refer to
  \sref{sNetwork} for an introduction into the network layer, and to
  \sref{sNetObject} for distributed objects.
\item[eq] The core namespace for the Equalizer client library. The class
  hierarchy in this namespace closely models the configuration file
  format. The classes in the \textsf{eq} namespace are the main subject
  of this Programming Guide.
\end{description}

>>>>

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications based on
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.6 RC1 for Equalizer 0.9\\[\medskipamount]
%  \today
}

@@

\lstset{language=C++}
\section*{Equalizer 0.9 Programming and User Guide}
\today
\subsection*{Contributors}

@@

OpenGL is a registered Trademark, OpenGL Multipipe is a Trademark of
Silicon Graphics, Inc. Linux is a registered Trademark of Linus
Torvalds.  Mac OS is a Trademark of Apple Inc. CAVELib is a registered
Trademark of the University of Illinois. The CAVE is a registered
Trademark of the Board of Trustees of the University of Illinois at
Chicago. Qt is a registered Trademark of Trolltech. TripleHead2Go is a
Trademark of Matrox Graphics. PowerWall is a Trademark of Mechdyne
Corporation. All other trademarks and copyrights herein are the property
of their respective owners.

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6 RC1 & Jul 31, 2009 & Release Candidate 1 for Equalizer 0.9\\
    \hline
  \end{tabularx}}
\clearpage

@@

Equalizer is the most advanced middleware for scalable 3D visualization,
providing the broadest set of parallel rendering features available in
an open source library to any OpenGL application. Many commercial and
open source applications in a variety of different markets rely on
Equalizer for flexibility and scalability.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
  {\caption{\label{fExecutionFlow}Parallel Rendering}}
  \vspace{-4ex}
\end{wrapfigure}

\fig{fExecutionFlow} illustrates the basic principle of any parallel
rendering application. The typical OpenGL application, for example GLUT,
has an event loop which redraws the scene, updates data based on
received events, and eventually redraws a new frame.

A parallel rendering application uses the same basic execution model and
extends it by separating the rendering code from the main event
loop. The rendering code is then executed in parallel on different
resources, depending on the configuration chosen at runtime.

This model is naturally followed by Equalizer, thus making application
development as easy as possible.

@@

The Equalizer architecture is based on a client-server model. The client
library exposes all functionality discussed in this document to the
programmer, and provides communication between the different Equalizer
processes. The underlying network layer uses a peer-to-peer structure,
allowing render clients to communicate directly with each other, e.g.,
during parallel compositing.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/processes.pdf}
  {\caption{\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The render client implements the rendering part of an application. Its
execution is passive, it has no main loop and is completely driven by
Equalizer. It executes the rendering tasks received from the server by
calling the appropriate task methods (see \sref{sTaskMethods}) in the
correct thread and context. The application either implements the task
methods with application-specific code or uses the default methods
provided by Equalizer.

The application can also be a rendering client, in which case it can
also contribute to the rendering. If it does not implement any render
client code, it is reduced to be the application's `master' process
without any OpenGL windows and 3D rendering.

@@

This section gives an introduction to scalable rendering, providing some
background for end users and application developers. The scalability
modes offered by Equalizer are discussed, along with their advantages
and disadvantages.

@@

2D decompositions can be used by all types of applications, but should
be combined with DB compounds to reduce the data per node, if
possible. In most cases, a \textsf{loadBalancer} should be used to
automatically adjust the tiling each frame, based on the current
rendering load.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\label{fPixel}\small A pixel compound}}
\end{wrapfigure}
OpenGL functionality influenced by the raster position will not work
correctly with pixel compounds, or needs at least special
attention. Among them are: lines, points, sprites, glDrawPixels,
glBitmap, glPolygonStipple. The application can query the current pixel
parameters at runtime to adjust the rendering accordingly.

Pixel compounds work well for purely fill-limited applications.
Techniques like frustum culling do not reduce the rendered data for the
source rendering resources. Pixel compounds are ideal for ray-tracing,
which is highly fill-limited and needs the full database for rendering
anyway. Volume rendering applications are also well suited for this
mode, and should choose it over 2D compounds.

@@

Some scalable rendering parameters are updated at runtime to provide
optimal performance or influence other scalable rendering
features. These adjustments are often referred to as load-balancing, but
are called \textsf{equalizers}, since their functionality is not limited
to load-balancing in Equalizer.

The server provides a number of equalizers, which automatically update
certain parameters of compounds based on runtime information. They
balance the load of 2D and DB compounds, optimally distribute render
resources for segment projections systems, adjust the resolution
to provide a constant framerate or zoom images to allow monitoring of
another view.

@@

\begin{itemize}
\item A list of all \textbf{computers} in your rendering cluster,
  including the \textbf{IP addresses} of all network interfaces to be
  used.
\item The number of \textbf{graphics cards} in each computer.
\item The \textbf{physical dimensions} of the display system, if
  applicable. These are typically the bottom-left, bottom-right and
  top-left corner points of each display surface in meters.
\item The relative coordinates of all the \textbf{segments} belonging to
  each display surface, and the graphics card \textbf{output} used for
  each segment. For homogenous setups, it is often enough to know the
  number of rows and columns on each surface, as well as the overlap or
  underlap percentage, if applicable.
\item The number of desired \textbf{application windows}. Application
  windows are typically destination windows for scalable rendering or
  'control' windows paired with a view on a display system.
\item \textbf{Characteristics} of the application, e.g., supported
  scalability modes and features.
\end{itemize}

@@

Equalizer applications are configured at runtime by the Equalizer
server. The server loads its configuration from a text file, which is a
one-to-one representation of the configuration data structures at
runtime.

@@

In each node, list all \textsf{connection}s through which this node is
reachable. Typically a node uses only one connection, but it is possible
to configure multiple connections if the machine and cluster is set up
to use multiple, independent network interfaces. Make sure the
configured \textsf{hostname} is reachable from all nodes. An IP address
may be used as the hostname.

@@

For each node, create on \textsf{pipe} for each graphics card in the
machine. Set the \textsf{device} number to the correct index. On
operating systems using X11, e.g., Linux, also set the port number if
your X-Server is running on a nonstandard port.

The \textsf{eq::Pipe} class is the abstraction of a graphics card (GPU).
In the current implementation it is also one operating system
thread. Non-threaded pipes are supported for integrating with
thread-unsafe libraries, but have various performance caveats. They
should only be used if using a different, synchronized rendering thread
is not an option.

@@

Configure one \textsf{window} for each desired application window on the
appNode. Configure one full-screen window for each display
segment. Configure one off-screen window, typically a pbuffer, for each
graphics card used as a source for scalable rendering. Provide a useful
name to each on-screen window if you want to easily identify it at runtime.

Sometimes display segments cover only a part of the graphics card
output. In this case it is advised to configure a non-fullscreen window
without window decorations, using the correct window viewport.

@@

Configure one \textsf{channel} for each desired rendering area in each
window. Typically one full-screen channel per window is used. Name the
channel using a unique, easily identifiable name, e.g., 'source-1',
'control-2' or 'segment-2\_3'.

@@

Sometimes, a single window is split across multiple projectors, e.g., by
using an external splitter such as the Matrox TripleHead2Go. In this
case configure one channel for each segment, using the channel's
\textsf{viewport} to configure its position relative to the window.

@@

A canvas represents one logical projection surface, e.g., a PowerWall, a
curved screen or an immersive installation. One configuration might
drive multiple canvases, for example an immersive installation and an
operator station.

A canvas consists of one or more segments. A planar canvas typically has
a frustum description (see \sref{sFrustum}), which is inherited by the
segments. Non-planar frusta are configured using the segment
frusta. These frusta typically describe a physically correct display
setup for Virtual Reality installations.

@@

For non-planar displays, configure the frustum as described in
\sref{sCanvas}. For passive stereo installations, configure one segment
per eye pass, where the segment for the left and right eye have the same
viewport.

A segment represents one output channel of the canvas, e.g., a projector
or an LCD. A segment has an output channel, which references the channel to
which the display device is connected.

A segment covers a part of its parent canvas, which is configured using
the segment viewport. The viewport is in normalized coordinates with
respect to the canvas. Segments might overlap (edge-blended projectors)
or have gaps between each other (display walls,
\fig{fCanvas}\footnote{Dataset courtesy of VolVis distribution of SUNY
  Stony Brook, NY, USA.}). The viewport is used to configure the
segment's default frustum from the canvas frustum description, and to
place layout views correctly.

@@

A view has a fractional viewport relative to its layout.  A layout
is often fully covered by its views, but this is not a requirement.

@@

Compound trees are used to describe how the rendering resources are
combined to produce the desired output, and how they are aggregated to
increase the performance.

It is advised to study and understand the basic configuration files
shipped with the Equalizer configuration, before attempting to write
compound configurations. The command line program \textsf{configTool},
shipped with the Equalizer distribution, creates some standard
configurations automatically.

For configurations using canvases and layouts without scalability, the
configuration file loader will create the appropriate compounds. It is
typically not necessary to write compounds for this use case.

The following subsection outlines the basic approach to writing
compounds. The remaining subsections provide an in-depth explanation of
the compound structure to give the necessary background for compound
configuration.

\subsubsection{Writing Compounds}
The following steps are typically taken when writing compound
configurations:
\begin{description}
\item [Root compound] Define an empty top-level compound when
  synchronizing multiple destination views. Multiple destination views
  are used for multi-display systems, e.g., a PowerWall or CAVE. All
  windows used for one display surface should be swap-locked (see below)
  to provide a seamless image. A single destination view is typically
  used for providing scalability to a single workstation window.
\item [Destination compound(s)] Define one compound for each destination
  channel, either as a child of the empty group, or as a top-level
  compound.
  \begin{description}
  \item[Standalone channels] Set the channel by using the channel name
    from the resource section. Define a frustum as explained in
    \sref{sCanvas}.
  \item[View/Segment channels] Set the channel by using the canvas,
    segment, layout and view name. The compound frustum will be
    calculated automatically based on the segment or view frustum. Note
    that one segment may created multiple view/segment channels, one for
    each view intersection of each layout used on the canvas. Only the
    compounds belonging to the active layout of a canvas are activate at
    runtime.
    \begin{description}
    \item [Software swap lock] When using software swap synchronization,
      swap-lock the destination compounds using a swapbarrier. All
      windows with a swapbarrier of the same name synchronize their
      swapbuffers.
    \item [Hardware swap lock] When using hardware swap synchronization,
      use swapbarriers for the destination compounds, setting
      \textsf{NV\_group} and \textsf{NV\_barrier} appropriately. The
      swap barrier name is ignored in this case. All windows of the same
      group on a single node synchronize their swap buffer. All groups
      of the same barrier synchronize their swap buffer across
      nodes. Please note that the driver typically limits the number of
      groups and barriers to one, and that multiple windows per swap
      group are not supported by all drivers. Please refer to your
      OpenGL driver documentation for details.
    \end{description}
  \end{description}
\item[Scalability] If desired, define scalability for each of your
  destination compounds. Add one compound using a source channel for
  each contributor to the rendering. The destination channel may also be
  used as a source. 
  \begin{description}
  \item[Decomposition] On each child compound, limit the rendering task
    of that child by setting the \textsf{viewport}, \textsf{range},
    \textsf{period} and \textsf{phase}, \textsf{pixel}, \textsf{eye} or
    \textsf{zoom} as desired.
  \item[Runtime Adjustments] A \textsf{load\_equalizer} may be used on
    the destination compounds to set the \textsf{viewport} or
    \textsf{range} of all children each frame, based on the current
    load. A \textsf{view\_equalizer} may be used on the root compound to
    assign resources to all destination compounds, which have to use
    \textsf{load\_equalizer}s. A \textsf{framerate\_equalizer} should be
    used to smoothen the framerate of DPlex compounds. A
    \textsf{DFR\_equalizer} may be used to set the zoom of a compound to
    achieve a constant framerate. One compound may have multiple
    equalizers, e.g., a \textsf{load\_equalizer} and a
    \textsf{DFR\_equalizer} for a 2D compound with a constant framerate.
  \item[Recomposition] For each source compound, define an
    \textsf{output\_frame} to read back the result. Use this output frame
    as an \textsf{input\_frame} on the destination compound. The frames
    are connected with each other by their name, which has to be unique
    within the root compound tree. For parallel compositing, describe
    your algorithm by defining multiple input and output frames across
    all source compounds.
  \end{description}
\end{description}

@@

Compounds have a frustum description to define the physical layout of
the display environment. The frustum specification is described in
\sref{sCanvas}.

@@

@@

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination channel's rendering, which is defined by the viewport,
frustum and database. A \textsf{viewport} decomposes the destination
channel and frustum in screen space. A \textsf{range} tells the
application to render a part of its database, and an \textsf{eye}
rendering pass can selectively render different stereo passes. A
\textsf{pixel} parameter adjusts the frustum so that the source channel
renders an even subset of the parent's pixels. Setting one or multiple
attributes causes the parent's view to be decomposed
accordingly. Attributes are cumulative, that is, intermediate compound
attributes affect and therefore decompose the rendering of all their
children.

@@

Note that a software swap barrier is not accurate enough for
edge-blended projection system or immersive installations. Such
installations typically use a hardware synchronization, e.g., nVidia
G-Sync cards. It is however sufficient for display walls made out of
LCD's and for some non-immersive projection systems.

@@

\subsubsection{\label{sEqualizers}Adjustments - Equalizers}

Equalizers are used to update compound parameters based on runtime
data. They are attached to a compound (sub-)tree, on which they
operate. The Equalizer distribution contains numerous example
configuration files using equalizers.

@@

Using a 2D or DB load-balancer will adjust the 2D split or database range
automatically each frame. The 2D load-balancer exists in three flavors:
\textsf{2D} using tiles, \textsf{VERTICAL} using columns and
\textsf{HORIZONTAL} using rows.

@@

DB load-balancing is beneficial for applications which cannot precisely
predict the load for their scene data, e.g., when the data is
nonuniform. Volume rendering is a counterexample, where the data is
uniform and a static DB decomposition typically results in a better
performance.

\paragraph{View Equalizer}
Depending on the model position and data structure, each segment of a
multi-display system has a different rendering load. The segment with
the biggest load determines the overall performance when using a static
assignment of resources to segments. The view equalizer analyzes the
load of all segments, and adjusts the resource usage each frame. It
equalizes the load on all segments of a view.

\fig{fViewLoadBalancing}\footnote{Image Copyright Realtime Technology
  AG, 2008} illustrates this process. On the left side, a static
assignment of resources to display segments is used. The right-hand
segment has a higher load than the left-hand segment, causing
sub-optimal performance. The configuration on the left uses a view
equalizer, which assigns two GPU's to the left segment and four GPU's to
the right segment, which leads to optimal performance for this model and
camera position.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
  {\caption{\label{fViewLoadBalancing}\small Cross-segment
      load-balancing for two segments using eight GPU's}}
\end{wrapfigure}
The view equalizer can also use resources from another display resource,
if this resource has little rendering load by itself. It is therefore
possible to improve the rendering performance of a multi-display system
without any additional resources. This is particularly useful for
installations with a higher number of displays where the rendering load
is typically in a few segments only, e.g., for a CAVE.

@@

\paragraph{DFR Equalizer}
Dynamic Frame Resolution (DFR) trades rendering performance for visual
quality. The rendering for a channel is done at a different resolution
than the native channel resolution to keep the framerate constant. The
DFR equalizer adjusts the zoom of a channel, based on the target and
current framerate. It is typically used for fill-rate bound
applications, such as volume rendering and ray-tracing.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
  {\caption{\label{fDFR}\small Dynamic Frame Resolution}}
\end{wrapfigure}
\fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
  Science Institute, University of Freiburg, Germany} shows DFR for
volume rendering. To achieve 10 frames per second, the model is rendered
at a lower resolution, and upscaled to the native resolution for
display. The rendering quality is slightly degraded, while the rendering
performance remains interactive. When the application is idle, it
renders a full-resolution view.

The dynamic frame resolution is not limited to downscaling the
rendering resolution, it will also upscale the resolution if the
parameters allow for it. Upscaled rendering, which will down-sample the
result for display, provides dynamic anti-aliasing at a constant
framerate.

\paragraph{Monitor Equalizer}
The monitor equalizer allows the observation of another view,
potentially made of multiple segments, in a different channel at a
different resolution. This is typically used to reuse the rendering of
a large-scale display on an operator station.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/monitorEq.pdf}
  {\caption{\label{fMonitorEq}\small Monitoring a projection wall}}
\end{wrapfigure}
A monitor equalizer adjusts the frame zoom of the output frames used to
observe the rendering, depending on the destination channel size. The
output frames are downscaled on the GPU before readback, which results
in optimal performance.

@@

To modify an application for Equalizer, the programmer structures the
source code so that the OpenGL rendering can be executed in parallel,
potentially using multiple processes for cluster-based execution.

@@

Methods called by the application have the form \textsf{verb[Noun]},
whereas methods called by Equalizer (`Task Methods') have the form
\textsf{nounVerb}. For example, the application calls
\textsf{Config::startFrame} to render a new frame, which causes, among
many other things, \textsf{Node::frameStart} to be called in all active
render clients.

@@

\begin{description}
\item[eq::base] The eq::base namespace provides C++ classes to abstract
  the underlying operating system and implements common helper
  functionality. One example is the \textsf{eq::base::Clock} class
  providing a high-resolution timer. Classes in this namespace are fully
  documented in the API documentation on the Equalizer website, and are
  not subject of this Programming Guide.
\item[eq::net] The Equalizer network layer provides basic functionality
  for network communication, such as \textsf{Connection} and
  \textsf{ConnectionSet}, as well as higher-level functionality such as
  \textsf{Node}, \textsf{Session} and \textsf{Object}. Please refer to
  \sref{sNetwork} for an introduction into the network layer, and to
  \sref{sNetObject} for distributed objects.
\item[eq] The core namespace for the Equalizer client library. The
  classes and their relationship in this namespace closely model the
  configuration file format. The classes in the \textsf{eq} namespace
  are the main subject of this Programming Guide.
\end{description}
